\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-06 paper.qmd
Added reasons for two clusters

The digit 1 of MNIST dataset consists of $7877$ grayscale images of handwritten digits [@lecun2010]. Before further analysis, PCA was used to preprocess the data, where the first $10$ principal components, explaining 83% of the total variation, were selected. The objective is to select the best NLDR layout, representing the structure of the digit 1 data in $10\text{-}D$ with default (hyper-)parameters.

@@

<!--PaCMAP param: n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0-->
```{r}
#| fig-cap: "$2\\text{-}D$ layout from PaCMAP applied for the digit 1 of the MNIST dataset. The (hyper-)parameter settings, beyond the defaults, are $10$ nearest neighbors, ratio of the number of mid-near pairs to the number of neighbors equal to $0.5$, and the ratio of the number of further pairs to the number of neighbors is $2$. We use our *model-in-the-data-space* to assess whether this is an accurate representation of non-linear structure present in $10\\text{-}D$, or if it is misleading. The angle of the digit 1 images varies along this structure. Images at the top-left of the $2\\text{-}D$ layout show the digit 1 angled more to the left, while those at the bottom-right show the digit 1 angled more to the right."
#| label: fig-pacmap-author
#| fig-pos: H
#| out-height: 22%

>>>>

The digit 1 of the MNIST dataset consists of $7877$ grayscale images of handwritten digits [@lecun2010]. Before further analysis, PCA was used to preprocess the data, where the first $10$ principal components, explaining 83% of the total variation, were selected. The objective is to select a reasonable \gD{} layout, representing the non-linear structure of the digit 1 dataset in $10\text{-}D$.

The MSE for different bin widths ($a_1$) using tSNE, UMAP, PHATE, PaCMAP, and TriMAP with default (hyper-)parameter setting (@fig-mnist-mse) were calculated. It is found that tSNE (@fig-mnist-mse (a)) provide the most reasonable representation for the digit 1 dataset, showing universally best. However, \gD{} representation shows a big non-linear cluster and a small cluster with a small gap.

In the case of digit 1 data, there should not be any clusters unless anomalies exist, indicating different digit 1 patterns. The angle of the digit 1 images varies along this non-linear clustering structure (@fig-tsne-best), while the small cluster contains the digit 1 images with different patterns of the digit 1, unlike the usual (@fig-model-error-mnist (c)).

@@

<!--PaCMAP param: n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0-->
```{r}
#| fig-cap: "The $2\\text{-}D$ layout from tSNE applied for the MNIST digit 1 dataset with defualt (hyper-)parameters. We use our *model-in-the-data-space* to assess whether this is an accurate representation of non-linear structure present in $10\\text{-}D$, or if it is misleading. The angle of the digit 1 images varies along this structure. Images at the top-right of the $\\gD{}$ layout show the digit 1 angled more to the right, while those at the bottom-right show the digit 1 angled more to the left."
#| label: fig-tsne-best
#| fig-pos: H
#| out-height: 22%

### 2024-09-06 paper.tex
Added reasons for two clusters

The digit 1 of MNIST dataset consists of \(7877\) grayscale images of
handwritten digits \citep{lecun2010}. Before further analysis, PCA was
used to preprocess the data, where the first \(10\) principal
components, explaining 83\% of the total variation, were selected. The
objective is to select the best NLDR layout, representing the structure
of the digit 1 data in \(10\text{-}D\) with default (hyper-)parameters.

@@

\includegraphics[width=1\textwidth,height=0.22\textheight]{paper_files/figure-pdf/fig-pacmap-author-1.pdf}

@@

\caption{\label{fig-pacmap-author}\(2\text{-}D\) layout from PaCMAP
applied for the digit 1 of the MNIST dataset. The (hyper-)parameter
settings, beyond the defaults, are \(10\) nearest neighbors, ratio of
the number of mid-near pairs to the number of neighbors equal to
\(0.5\), and the ratio of the number of further pairs to the number of
neighbors is \(2\). We use our \emph{model-in-the-data-space} to assess
whether this is an accurate representation of non-linear structure
present in \(10\text{-}D\), or if it is misleading. The angle of the
digit 1 images varies along this structure. Images at the top-left of
the \(2\text{-}D\) layout show the digit 1 angled more to the left,
while those at the bottom-right show the digit 1 angled more to the
right.}

@@

\caption{\label{fig-mnist-tri-proj}Model in \gD{}, on the tSNE layout
(a), and three views (b, c, d) of the fit in projections from
\(10\text{-}D\), for the digit 1 of MNIST data (\(a_1 = 0.062\),
\(b = 285 \  (19,  15)\), \(m = 109\), and benchmark value to remove
large edges is \(0.108\)). (b) The non-linear structure observed in the
\gD{} layout of PaCMAP (Figure~\ref{fig-pacmap-author}) is also visible
when visualizing the model overlaid on the data space. This indicates
that PaCMAP accurately captures the non-linear structure of the
\(10\text{-}D\) data. (c) the model shows a twisted pattern within the
non-linear structure in \(10\text{-}D\) space, which is an additional
pattern not visible in the \gD{} representation
(Figure~\ref{fig-pacmap-author}). (d) some long edges exist in the
\(10\text{-}D\) space that are not recognized as long edges in the \gD{}
representation. The full video is available at
\url{https://youtu.be/zcg_GXBmqjA}.}

>>>>

The digit 1 of the MNIST dataset consists of \(7877\) grayscale images
of handwritten digits \citep{lecun2010}. Before further analysis, PCA
was used to preprocess the data, where the first \(10\) principal
components, explaining 83\% of the total variation, were selected. The
objective is to select a reasonable \gD{} layout, representing the
non-linear structure of the digit 1 dataset in \(10\text{-}D\).

The MSE for different bin widths (\(a_1\)) using tSNE, UMAP, PHATE,
PaCMAP, and TriMAP with default (hyper-)parameter setting
(Figure~\ref{fig-mnist-mse}) were calculated. It is found that tSNE
(Figure~\ref{fig-mnist-mse} (a)) provide the most reasonable
representation for the digit 1 dataset, showing universally best.
However, \gD{} representation shows a big non-linear cluster and a small
cluster with a small gap.

In the case of digit 1 data, there should not be any clusters unless
anomalies exist, indicating different digit 1 patterns. The angle of the
digit 1 images varies along this non-linear clustering structure
(Figure~\ref{fig-tsne-best}), while the small cluster contains the digit
1 images with different patterns of the digit 1, unlike the usual
(Figure~\ref{fig-model-error-mnist} (c)).

@@

\includegraphics[width=1\textwidth,height=0.22\textheight]{paper_files/figure-pdf/fig-tsne-best-1.pdf}

@@

\caption{\label{fig-tsne-best}The \(2\text{-}D\) layout from tSNE
applied for the MNIST digit 1 dataset with defualt (hyper-)parameters.
We use our \emph{model-in-the-data-space} to assess whether this is an
accurate representation of non-linear structure present in
\(10\text{-}D\), or if it is misleading. The angle of the digit 1 images
varies along this structure. Images at the top-right of the \(\gD{}\)
layout show the digit 1 angled more to the right, while those at the
bottom-right show the digit 1 angled more to the left.}

@@

\caption{\label{fig-mnist-tri-proj}Model in \gD{}, on the tSNE layout
(a), and three views (b, c, d) of the fit in projections from
\(10\text{-}D\), for the digit 1 of MNIST data (\(a_1 = 0.062\),
\(b = 285 \  (19,  15)\), \(m = 109\), and benchmark value to remove
large edges is \(0.108\)). (b) The non-linear structure observed in the
\gD{} layout of PaCMAP (\textbf{?@fig-pacmap-author}) is also visible
when visualizing the model overlaid on the data space. This indicates
that PaCMAP accurately captures the non-linear structure of the
\(10\text{-}D\) data. (c) the model shows a twisted pattern within the
non-linear structure in \(10\text{-}D\) space, which is an additional
pattern not visible in the \gD{} representation
(\textbf{?@fig-pacmap-author}). (d) some long edges exist in the
\(10\text{-}D\) space that are not recognized as long edges in the \gD{}
representation. The full video is available at
\url{https://youtu.be/zcg_GXBmqjA}.}
